An Event-based Architecture for Multi-population Optimization Algorithms

### 2020-01-26 draft.tex
A few minor corrections

The algorithm starts with the Setup process pulling a configuration message from
the experiements queue, this is not shown in the diagram. The configuration messsage 
includes all the parameters needed to execute the algorithm, among other things,
the number of subpopulations, the number of individuals, and the number of iterations
of the algorithm. We give more details about the configuration data structure
in Section \ref{experiment_flow}. % Missing  % added 
Once the configuration is read, we can follow the path of 
messagess as follows:

\begin{enumerate}
\item In this step, the specified number of subpopulations is created acording to the 
parameters found in the configuration structure. The population at this moment is just
static data, including each individual inside. Each subpopulation includes a metadata
section where its algorithm and execution parameters are specified. For instance, 
for a GA, the mutation rate, type of crossover operator, and other values are indicated.

@@

\item Once they finish the execution, the population state
is packed again along with aditional metadata about the execution of the algorithm.
The resulting populations are now pushed to the Output queue. Once finished, another 
subpopulation is pulled from the queue.

@@

This reactive architecture has the following advantages:\begin{itemize}
\item 
An important aspect of proposal is the decoupling of the population and 
the population-based algorithm. It is common that in a classic island-based algorithm
each island is executed in a separate processing node, i.e. in a virtual machine, cpu,
or thread. In this case we can have just one processing node, or many nodes, each 
running a different search strategy, or using its own parameters. 
\item 
Also, the reactive controller gives designers more control over the multi-population algorithm.
In this process, designers can dynamically change the number of populations, population parameters,
and migration details on-the-fly.
\item 
Another advantege is that algorithm designers have many options for implementing this simple architecture. 
The same basic components can be implmented as a single multithreaded program, or 
as a highly scalable serverless cloud application. Most modern languages include 
constructs for asynchronous programming using queues or channels, for a multi-threaded
execution.
\end{itemize}

@@

On the other hand, there are several caveats as well:
It is more costly to move entire subpopulations as messages than only passing certain individuals 
from a process to another. Designers must conmsider this cost when working with large
individuals, and if possible send sub-populations using several messages, use compression or 
adjust the size of subpopulations. Pool-based algorithms also suffer from this drawback, but 
web based immplementations have been working with continous optimization problems of 1000 dimensions.
But, in other cases this is not a viable approach.

@@

Next, we compare the message-driven architecture we presented against four other
proposals for multi-population algorithms found in the literature. We center the
comparison on the coupling and communication between the main components:
Subpopulations, Processing Nodes, and Algorithms.  In Figure \ref{fig:population_message} we show the
main components of a message-driven architecture. We can see that the algorithm
and the subpopulations are separated, and the Population Queue keeps the state
of subpopulations. While algorithms only need the subpopulations as parameters
for their execution.

@@

We show another pool-based approach in Figure  \ref{fig:population_pool}. In
this design, the global population is stored in a central repository, while
isolated algorithms take random samples of the global population and use this
temporal subpopulation as parameters. An advantage of this design is that the
sampling provides a type of migration between isolated subpopulations. A problem
found is that when a processing node returns a subpopulation to the pool, the
state of the subpopulation is lost. But having the algorithm decoupled opens the
possibility of having a serverless-function implementation.

@@

The controller is an essential component of the architecture (Figure \ref{fig:kafkEO})
because it is responsible for maintaining the evolutionary loop. It takes newly
evolved populations from the Output Queue, mixes them,   and produces new
messages from the result. At the same time, it must keep track of two conditions
for ending the loop: the number of messages it has received reaches a maximum
value, or the error of the best solution goes below a specific threshold.  Finally, it
must filter out remnant messages from other experients. Messages from other
experiments can remain in the queues because of the asynchronous nature of the
system.  We chose to implement the controller in Python but using an API
specialized in asynchronous event-driven programming over data streams.  
The open-source library is called Reactive Extensions (ReactiveX) for Python 
\url{https://github.com/ReactiveX/RxPY} an it is based on the Observer 
pattern \cite{gamma1995design} and functional programming.

>>>>

The algorithm starts with the Setup process pulling a configuration message from
the experiments queue, this is not shown in the diagram. The configuration message 
includes all the parameters needed to execute the algorithm, among other things,
the number of subpopulations, the number of individuals, and the number of iterations
of the algorithm. We give more details about the configuration data structure
in Section \ref{experiment_flow}. % Missing  % added 
Once the configuration is read, we can follow the path of 
messagess as follows:

\begin{enumerate}
\item In this step, the specified number of subpopulations is created according to the 
parameters found in the configuration structure. The population at this moment is just
static data, including each individual inside. Each subpopulation includes a metadata
section where its algorithm and execution parameters are specified. For instance, 
for a GA, the mutation rate, type of crossover operator, and other values are indicated.

@@

\item Once they finish the execution, the population state
is packed again along with additional metadata about the execution of the algorithm.
The resulting populations are now pushed to the Output queue. Once finished, another 
subpopulation is pulled from the queue.

@@

This reactive architecture has the following advantages:\begin{itemize}
\item 
An important aspect of proposal is the decoupling of the population and 
the population-based algorithm. It is common that in a classic island-based algorithm
each island is executed in a separate processing node, i.e. in a virtual machine, cpu,
or thread. In this case we can have just one processing node, or many nodes, each 
running a different search strategy, or using its own parameters. 
\item 
Also, the reactive controller gives designers more control over the multi-population algorithm.
In this process, designers can dynamically change the number of populations, population parameters,
and migration details on-the-fly.
\item 
Another advantage is that algorithm designers have many options for implementing this simple architecture. 
The same basic components can be implmented as a single multi-threaded program, or 
as a highly scalable serverless cloud application. Most modern languages include 
constructs for asynchronous programming using queues or channels, for a multi-threaded
execution.
\end{itemize}

@@

On the other hand, there are several caveats as well:
It is more costly to move entire subpopulations as messages than only passing certain individuals 
from a process to another. Designers must consider this cost when working with large
individuals, and if possible send sub-populations using several messages, use compression or 
adjust the size of subpopulations. Pool-based algorithms also suffer from this drawback, but 
web based implementations have been working with continuous optimization problems of 1000 dimensions.
But, in other cases this is not a viable approach.

@@

Next, we compare the message-driven architecture we presented against four other
proposals for multi-population algorithms found in the literature. We center the
comparison on the coupling and communication between the main components:
Subpopulations, Processing Nodes, and Algorithms.  In Figure \ref{fig:population_message} we show the
main components of a message-driven architecture. We can see that the algorithm
and subpopulations are separated, and the Population Queue keeps the state
of subpopulations. While algorithms only need the subpopulations as parameters
for their execution.

@@

We show another pool-based approach in Figure  \ref{fig:population_pool}. In
this design, the global population is stored in a central repository, while
isolated algorithms take random samples of the global population and use this
temporal subpopulation as parameters. An advantage of this design is that the
sampling provides a type of migration between isolated subpopulations. A problem
found is that when a processing node returns a subpopulation to the pool, the
state of the subpopulation is lost. But having the algorithm decoupled opens the
possibility of a serverless-function implementation.

@@

The controller is an essential component of the architecture (Figure \ref{fig:kafkEO})
because it is responsible for maintaining the evolutionary loop. It takes newly
evolved populations from the Output Queue, mixes them,   and produces new
messages from the result. At the same time, it must keep track of two conditions
for ending the loop: the number of messages it has received reaches a maximum
value, or the error of the best solution goes below a specific threshold.  Finally, it
must filter out remnant messages from other experiments. Messages from other
experiments can remain in the queues because of the asynchronous nature of the
system.  We chose to implement the controller in Python but using an API
specialized in asynchronous event-driven programming over data streams.  
The open-source library is called Reactive Extensions (ReactiveX) for Python 
\url{https://github.com/ReactiveX/RxPY} an it is based on the Observer 
pattern \cite{gamma1995design} and functional programming.
